---
title: "P8105 Homework 3"
author: Pavithra Srinivasan, 
output: github_document
date: "11 Oct 2023"
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)

library(p8105.datasets)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

### Problem 1
#### Read in the data

```{r}
data("instacart")

instacart = 
  instacart |> 
  as_tibble()
```

#### Answer questions about the data

This dataset contains `r nrow(instacart)` rows and `r ncol(instacart)` columns, with each row resprenting a single product from an instacart order. Variables include identifiers for user, order, and product; the order in which each product was added to the cart. There are several order-level variables, describing the day and time of the order, and number of days since prior order. Then there are several item-specific variables, describing the product name (e.g. Yogurt, Avocado), department (e.g. dairy and eggs, produce), and aisle (e.g. yogurt, fresh fruits), and whether the item has been ordered by this user in the past. In total, there are `r instacart |> select(product_id) |> distinct() |> count()` products found in `r instacart |> select(user_id, order_id) |> distinct() |> count()` orders from `r instacart |> select(user_id) |> distinct() |> count()` distinct users.

Below is a table summarizing the number of items ordered from aisle. In total, there are 134 aisles, with fresh vegetables and fresh fruits holding the most items ordered by far.

```{r}
instacart |> 
  count(aisle) |> 
  arrange(desc(n))
```

Next is a plot that shows the number of items ordered in each aisle. Here, aisles are ordered by ascending number of items.

```{r}
instacart |> 
  count(aisle) |> 
  filter(n > 10000) |> 
  mutate(aisle = fct_reorder(aisle, n)) |> 
  ggplot(aes(x = aisle, y = n)) + 
  geom_point() + 
  labs(title = "Number of items ordered in each aisle") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

```

Our next table shows the three most popular items in aisles `baking ingredients`, `dog food care`, and `packaged vegetables fruits`, and includes the number of times each item is ordered in your table.

```{r}
instacart |> 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |>
  group_by(aisle) |> 
  count(product_name) |> 
  mutate(rank = min_rank(desc(n))) |> 
  filter(rank < 4) |> 
  arrange(desc(n)) |>
  knitr::kable()
```

Finally is a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week. This table has been formatted in an untidy manner for human readers. Pink Lady Apples are generally purchased slightly earlier in the day than Coffee Ice Cream, with the exception of day 5.

```{r}
instacart |>
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>
  group_by(product_name, order_dow) |>
  summarize(mean_hour = mean(order_hour_of_day)) |>
  pivot_wider(
    names_from = order_dow, 
    values_from = mean_hour) |>
  knitr::kable(digits = 2)
```

## Problem 2 
```{r}
library(p8105.datasets)
data("brfss_smart2010")
```

#### First, do some data cleaning:
- format the data to use appropriate variable names;
- focus on the “Overall Health” topic
- include only responses from “Excellent” to “Poor”
- organize responses as a factor taking levels ordered from “Poor” to “Excellent”

```{r}
brfss_df = 
  brfss_smart2010 |>
  janitor::clean_names() |>
  select(-location_id) |>
  filter(topic == "Overall Health") |>
  mutate(
    response = factor(response, levels = c("Poor", "Fair", "Good", "Very good", "Excellent"))
  ) |>
  arrange(response)
```
Data set description: There are a total of `r length(brfss_df)` variables,  `r nrow(brfss_df)` rows and `r ncol(brfss_df)` columns.

#### Answer questions about the data:

##### 1. In 2002, which states were observed at 7 or more locations? What about in 2010?

```{r}
State_Location_2002 = 
  brfss_df |>
  filter(year == 2002) |>
  group_by(year, locationabbr) |>
  summarize(nloc = n_distinct(geo_location)) |>
    filter(nloc >= 7) |>
    knitr::kable()

State_Location_2010 = 
  brfss_df |>
  filter(year == 2010) |>
  group_by(year, locationabbr) |>
  summarize(nloc = n_distinct(geo_location)) |>
    filter(nloc >= 7) |>
    knitr::kable()
```
In 2002,  6 states (CT, FL, MA, NC, NJ, PA) were observed at 7 or more locations and in 2010, 14 states (CA, CO, FL, MA, MD, NC, NE, NJ, NY, OH, PA, SC, TX, WA) were observed at 7 or more location.


##### 2. Construct a dataset that is limited to Excellent responses, and contains, year, state, and a variable that averages the data_value across locations within a state. Make a “spaghetti” plot of this average value over time within a state (that is, make a plot showing a line for each state across years – the geom_line geometry and group aesthetic will help).

```{r}
excellent_responses =
  brfss_df |>
  filter(response == "Excellent") |>
  select(year, locationabbr, data_value) |>
  group_by(locationabbr, year) |>
  mutate(
    avg_data_value = mean(data_value, na.rm = TRUE)
  ) 
spaghetti_plot =
  ggplot(excellent_responses, aes(x = year, y = avg_data_value, group = locationabbr, color = locationabbr)) +
  geom_line() +
  labs(
    title = "Avg data value within a state",
    x = "Year",
    y = "Avg data value",
    color = "State",
    theme(plot.title = element_text(hjust = 0.5)
    ))
spaghetti_plot
```

Observations: The Avg Data Value seems to fluctuate between 2002-2010 for most states, but notably there was a steep decrease in West Virginia in the year 2005 and again in 2009. 


##### 3. Make a two-panel plot showing, for the years 2006, and 2010, distribution of data_value for responses (“Poor” to “Excellent”) among locations in NY State

```{r}
brfss_df |> 
  select(year, data_value, locationabbr, locationdesc, response) |> 
  filter(
    year == 2006 | year == 2010,
    locationabbr == "NY") |> 
  ggplot(aes(x = data_value, fill = response)) +
  geom_boxplot(aes(fill = response), alpha = .5) +
   
  labs(
    title = "Data Values Across different Locations in NY State",
    x = "Data Value",
    fill = "Response") + 
  facet_grid(. ~ year)
```

Observations: Overall, 2006 and 2010 have a similar distribution of data values (ranging "excellent" to "poor") across NY State locations. Data values were the lowest among those who answered "poor" and the highest among those who answered "very good." The plots also look like it can potentially be left skewed, and there were more data values recorded in 2010. 

## Problem 3

##### Read, tidy, merge, and organize the nhanes accel and covar data sets. Your final dataset should include all originally observed variables; exclude participants less than 21 years of age, and those with missing demographic data; and encode data with reasonable variable classes


```{r}
accel_df = 
  read_csv("./data/nhanes_accel.csv") |>
  janitor::clean_names()
  
  
covar_df =
  read_csv("./data/nhanes_covar.csv", skip = 4) |>
  janitor::clean_names() |> 
  mutate(
    sex = case_match(
      sex,
      1 ~ "Male",
      2 ~ "Female"),
    education = case_match(
      education,
      1 ~ "Less than High School",
      2 ~ "High School Equivalent",
      3 ~ "More than High School"),
    education = factor(
      education,
      levels = c("Less than High School", "High School Equivalent", "More than High School")
    ))

nhanes_combined = 
  full_join(accel_df, covar_df) |> 
  drop_na() |> 
  filter(age >= 21)

view(nhanes_combined)
```

##### Produce a reader-friendly table for the number of men and women in each education category, and create a visualization of the age distributions for men and women.

```{r}
nhanes_education_table = 
  nhanes_combined |>
  group_by(sex,education) |>
  summarize(count = n()) |>
  knitr::kable(digits = 1)
```

Lets create a plot for the combined data!

```{r}
nhanes_education_table = 
  ggplot(nhanes_combined, aes(x = education, y = age, fill = sex)) +
  geom_boxplot() +
  labs(title = "Distribution of Age by Gender Across Education Levels",
       x = "Education Level",
       y = "Age") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

nhanes_education_table
```

Description: The plot shows the age distributions by gender in each education category. Both men and women with education levels above high school have a lower avg age compared to those in other education categories. Specifically, among the female group those who had a high school education had the highest density of people, versus in the male group those with below high school education levels had a higher density of people. In terms of age, women with a high school education were older than men at the same educational level. 


##### Using your tidied dataset, aggregate across minutes to create a total activity variable for each participant. Plot these total activities (y-axis) against age (x-axis); your plot should compare men to women and have separate panels for each education level. Include a trend line or a smooth to illustrate differences. Comment on your plot.

```{r}
nhanes_total_activity_plot = 
  nhanes_combined |> 
  group_by(education, sex) |> 
  mutate(total_activity = rowSums(across(min1:min1440))) |> 
  ggplot(aes(x = age, y = total_activity, color = education)) +
  geom_point(alpha = .5) +
  facet_wrap(~education) +
  geom_smooth() +
  labs(
    title = "Total Activity by Age Across Education Levels",
    x = "Age",
    y = "Activity") +
  theme(plot.title = element_text(hjust = 0.5))

nhanes_total_activity_plot
```
Description: There is a direct proportional relationship between age and activity. As participants age, their total level of activities decreases with time. Similarly, there is a decrease in activity across all educational levels as well. One thing to note, is that participants in the "less than high school" education category had the highest total activity at 20 years of age and also the lowest activity at 80 years of age. 

##### Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex. 

```{r}
three_panel_plot = 
  nhanes_combined |> 
  group_by(education, sex) |>
  summarise(across(starts_with("min"), ~ mean(.), .names = "mean_{.col}")) |>
  pivot_longer(cols = starts_with("mean_"), names_to = "time", values_to = "mean_value") |>
  mutate(time = as.numeric(substring(time, 9))) |> 
  ggplot(aes(x = time, y = mean_value, color = sex)) +
  geom_point() + 
  geom_smooth() +
  facet_grid(. ~ education) +
  labs(x = "Time (min)", y = "Value") +
  ggtitle("Activity over 24 hours by gender and education levels") +
  theme_minimal()

three_panel_plot
```

Observations: The distribution of MIMS is similar among males and females. Overall the "less than high school education" category has the highest mean value of MIMS. Male have a lower value of MIMS than female across all education levels. MIMS values start decreasing steadily after approx 1200min across all education levels and gender. 
